### 2025-11-09 thesis/implementation/index.tex
Challenges

>>>>

\subsection{Engineering Challenges and Lessons Learned}

The implementation approach of \textsc{Natural} highlighted significant
challenges in terms of practicability that were independent of the algorithmic
design and theoretical problems. While \textsc{Natural} turned out to
demonstrate the approaches recommended in the system design section, the
development process was significantly slowed down due to constrained hardware
availability, technology stack decisions and research methodology. The overall
system architecture turned out positive, but the timeline was noticably expanded.

This section reflects on the most impactful challenges that were encountered
during the implementation of \textsc{Natural} and analyzes how hardware
limitations influenced design decisions, confidence in the performance of
\textsc{Natural} and the overall impact on development velocity. These insights
are transferrable beyond \textsc{Natural} and are applicable to any
production-grade implementation of research algorithms.

\subsubsection{Hardware Constraints and Development Velocity}

Likely the most significant circumstance that affected development velocity was
the constrained access to high performance hardware. The hardware available
during development time was the RTX 3090 with 24GB of VRAM. This limitation
fundamentally shaped the development process and research methodology as the
iteration speed was severly slowed down.

\paragraph{Benchmarking and Evaluation Bottlenecks}

A full benchmark execution of \textsc{Natural} against the \textsc{Spider} test
dataset required around \textcolor{red}{12-36 hours} depending on the number of
candidates generated and refinement algorithm used during the benchmark. 
This made continuous validation during development a non trivial task,
requiring extensive evaluation phases between times of active development.

Testing new hypotheses, validating algorithmic changes, experimenting with the
implementation approaches, comparing different configurations and sampling
strategies therefore became a multi-week processes rather than the rapid
iteration typically desired in research environments.

The prolonged evaluation cycles created a cascading effect on development
velocity. Rather simple adjustments that would normally be validated within
hours required days of computational time, effectively preventing what should
have been normal experimentation. The hardware constraints forced a rather
conservative approach to experimentation, where each change needed to be
carefully considered before committing to multiple days of benchmarking.

\paragraph{Memory Management and CUDA Issues}

Frequent CUDA out-of-memory (OOM) errors made the evaluation interpretation non
trivial as accuracy scores became inaccurate due to pipeline failures. This persistet as a

second development challenge throughout the implementation phase. These memory exhaustion issues
occurred rather unpredictable during both development and benchmarking,
depending on the database schema and question asked. As \textsc{Natural} uses a
code representation prompt, the database schema is inlined into the prompt
during inference. If \textsc{Natural} is executed against large databases, even
with schema subsetting, the context window can be exceeded or the KV-Cache can
exhaust the rest of the available GPU memory (typically around 2-4 gigabytes)
after models have been loaded. This required extensive trial-and-error
debugging to identify the root causes and implement workarounds.

\subsubsection{Technology Stack Trade-offs}

During the implementation of \textsc{Natural}, several critical technology
stack decisions fundamentally impacted both development velocity and the
final system characteristics. The two most consequential decisions were the
choice of programming language and the model deployment strategy, each
presenting distinct trade-offs between research efficiency and production
readiness.

\paragraph{Programming Language: Rust vs Python}

The decision to implement \textsc{Natural} in Rust rather than Python
represents perhaps the most impactful architectural choice made during the
implementation phase. Choosing Rust presented the opportunity to port the
research system into production environments soon after, but came at the cost
of having to port existing benchmarking infrastructure and not being able to
exactly reuse the inference code used by \textsc{OmniSQL}. This decision
significantly increased implementation time, likely doubling the development
effort compared to what would have been required in Python.

Using Python would have provided immediate access to existing machine
learning infrastructure and could have leveraged existing implementations by
previous researchers directly in the implementation of \textsc{Natural}.

The rather mature Python ecosystem for NLP and ML would have eliminated the
need to develop abstractions for model inference, vector database operations,
and benchmark evaluation.

However, the Rust implementation as is provides significant benefits for production
deployment and system integration. The resulting system can be embedded
directly into database systems such as PostgreSQL through extensions,
providing a path toward true production deployment that would be impractical
with Python. The ability to compile an entire NL2SQL system into a single,
standalone binary offers superior portability characteristics.

\paragraph{Model Deployment: Local vs Cloud-Based Inference}

Using local and open source models reemphasized the limited local hardware
access as using a cloud provider like OpenAI or Anthropic for inference would
have singificantly speeded up the benchmarking time. As a primary research goal
of this thesis is to explore the open source capabilities of NL2SQL systems,
choosing a proprietary inference service was deeemed unviable.

\paragraph{Integration Complexity and Ecosystem Maturity}

The machine learning ecosystem of Rust proved to be singificantly less mature
than anticipated, requiring extensive custom development and research for
functionality that would be available off-the-shelf in prevalent Python packages.
The integration with Python libraries for specialized components like the
Wasserstein-Weisfeiler-Leman kernels required rather complex FFI code using
\texttt{pyo3}, further adding architectural complexity and potential stability
concerns.

Despite these challenges, the resulting architecture demonstrates that multi
language approaches can be viable when different components have distinct
requirements. The Python integration was isolated to the sampling phase,
preserving the runtime portability characteristics of the core Rust
implementation while still leveraging existing, specialized libraries where
appropriate.
